\section{Formalization}
\label{sec:formalization}

\subsection{Operational Semantics}
\label{sec:opsem}

We formalize our ideas in the context of a lambda calculus ($\lang$)
shown in Fig.~\ref{fig:opsem}. Expressions of $\lang$ are variables,
constants, and \name primitives composed using the lambda combinator.
For brevity, we use short names for \name primitives: \C{run} for
\C{with\_init\_version\_do}, and \C{fork} for \C{fork\_version}. To
simplify the technical development, \name's \C{sync\_next\_version}
operation is broken down into two primitives - \C{push} and \C{pull},
which can be composed to get the desired effect of the original:
$\C{sync}\;x \;=\; (\lambda y.\pull)\; (\push\,x)$.  The semantics
of\C{get\_current\_version} is subsumed by \C{pull}, hence elided.
Values ($v$) are constants and lambda abstractions.  A program ($p$)
is a parallel composition of threads, where each thread is an
expression ($s$) indexed by the corresponding thread identifier ($t$).

\input{opsem}

Fig.~\ref{fig:opsem} also shows the syntax of \emph{branches}, which
are artifacts of evaluation and only appear during run-time. A branch
is a non-empty sequence of tagged values, where the tag captures the
abstract run-time operation that led to the creation of the value. It
is implicitly assumed that each value added to a branch is uniquely
identifiable, and hence no two values on a branch are equal.  The
uniqueness assumption is later extended to a collection of branches
that constitute a branching structure. A real implementation meets
this assumption by versioning values across the branches. Thus, in
reality, branches contain \emph{versions} which denote values.  To
simplify the presentation, the semantics, does not make this
distinction, and uses values and versions interchangeably.

The small-step operational semantics of $\lang$ is defined via a
reduction relation ($\stepsto$) that relates \emph{program states}. A
program state ($p;\,H$) consists of a program $p$ and a \emph{branch
history} $H$ that maps thread identifiers to corresponding branches.
Each thread is associated with a branch during evaluation. Evaluation
contexts have been defined separately for expressions ($E$) and
programs ($P$), with the latter subsuming the former.

$E$ is defined to evaluate the first argument of a \C{run} expression
to a value that constitutes an initial version (recall that \C{run}
models \nameMonad's \C{with\_init\_version\_do}). A program evaluation
context non-deterministically picks an extant thread to evaluate. The
administrative rule that relates transitions of holes to transitions
of expressions and programs is straightforward, and hence elided. The
remaining core reduction rules are presented in
Fig.~\ref{fig:opsem}. For brevity, we write $H(t\mapsto (v,f))$ to
denote the proposition that $H$ maps $t$ to $(v,f)$. The notation $H[t
  \mapsto (v,f)]$ denotes the extension of $H$ with the binding $t
\mapsto (v,f)$, as usual.

Reduction rules let expression evaluation take a step by rewriting the
expression and suitably updating the branch history ($H$).
\rulelabel{E-App} is the standard beta reduction rule.  The
\rulelabel{E-Run} rule is applicable only when $H$ is empty, i.e.,
when no prior branching structure exists, and commences a distributed
computation with a corresponding version tree. The rule rewrites the
$\C{run}\;v\;s$ expression to $s$, while creating a new branching
structure with two branches: a \emph{top} branch that has just the
initial version (tagged with \C{INIT}), and a branch for the current
thread ($t$) forked-off from the top branch.  The first version on the
current branch ($H(t)$) denotes the same value ($v$) as the initial
version on the top branch, although versions themselves are deemed
distinct. The new version is tagged with a \C{FORK} tag that keeps the
record of its origin, namely the \C{fork} operation and the branch from
which the current branch is forked. The \rulelabel{E-Fork} rule forks
a new thread with a fresh id ($t'$) and adds it to the thread pool.
The corresponding branch ($H(t')$) is forked from the parent thread's
branch ($H(t)$). The semantics of branch forking is the same as
described above. The \C{fork} expression in the parent thread
evaluates to \C{()}. The \rulelabel{E-Push} rule creates a new version
on the current branch ($H(t)$) using the pushed value ($v$).  Although
our semantics does not directly expose heap-allocated values, the
intention is that $v$ is a replicated object, manipulated on the local
heap that, after push, now becomes subject to merging and coordination
with other replicas.

The semantics non-deterministically chooses \rulelabel{E-Pull} or
\rulelabel{E-Pull-Wait} rules to reduce a \C{pull} expression. The
\rulelabel{E-Pull} rule reduces \C{pull} to \C{()}, and returns the
latest version on the current branch. The \rulelabel{E-Pull-Wait} rule
can be thought of as a stutter step; it doesn't reduce \C{pull}, but
updates the branching structure by merging (the latest version of) a
concurrent branch ($H(t')$) into (the latest version of) the current
branch ($H(t)$), and extending the current branch with the merged
version ($v_m$). The versions are merged only if they are \emph{safely
mergeable} (denoted $\under{H}{v' \mbleto v}$) - a notion that we will
explain shorty. The new version is tagged with a \C{MERGE} tag that,
like a \C{FORK} tag, records its origin. The rule assumes a function
\C{lca} that computes the \emph{lowest common ancestor} (LCA) for the
latest versions on the given pair of branches. The formal definition
of LCA is given below. The \rulelabel{E-Pull-Wait} and
\rulelabel{E-Pull} rules thus let a thread synchronize with other
distributed threads manipulating versions of replicas in multiple
steps before returning the result of the \C{pull}. Since \C{sync} is a
composition of \C{push} and \C{pull}, its behavior can be explained
thus: \C{sync} pushes the given value onto the current (local) branch,
merges a (possibly empty) subset of concurrent branches into the local
branch, and returns the result.  (The operation is not guaranteed to
synchronize with all concurrent branches because not all such branches
may be available, because of e.g., network partitions.)

To define LCA, we formalize the intuitive notation of the ancestor
relationship between versions of a legal branching history (i.e., a
branching history generated by the rules in Fig.~\ref{fig:opsem}):

\begin{definition} [\bfseries Ancestor]
Version $v_1$ is a ancestor of version $v_2$ under a history
$H$ (written $\under{H}{v_1 \preceq v_2}$) if and only if one of the
following is true:
\begin{itemize}
  \item There exists a branch $b$ in $H$ (i.e., $\exists(t\in
  dom(H)).\,H(t) = b$) in which $v_2$ immediately succeeds
  $v_1$,
  \item There exists a branch $b$ in $H$ that contains $(v_2,
  \C{FORK}\; (v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
  \item There exists a branch $b$ in $H$ that contains
  $(v_2, \C{MERGE}\;(v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
  \item $v_1 = v_2$, or $v_1$ is transitively an ancestor of
  $v_2$, i.e., $\exists v.~ \under{H}{v_1 \preceq v} \conj
  \under{H}{v \preceq v}$
\end{itemize}
\end{definition}

The ancestor relation is therefore a partial order with a greatest
lower bound (the initial version).  Thus, for any two versions in a
legal history, there exists at least one common ancestor.  The ancestor
relationship among all common ancestors lets us define the notion of a
lowest common ancestor (LCA):

\begin{definition} [\bfseries Lowest Common Ancestor]
Version $v$ is said to be a common ancestor of versions $v_1$ and
$v_2$ under a history $H$ if and only if $\under{H}{v \preceq v_1}$
and $\under{H}{v \preceq v_2}$. It is said to be the lowest common
ancestor (LCA) of $v_1$ and $v_2$, iff there does not exist a $v'$
such that $\under{H}{v' \preceq v_1}$ and $\under{H}{v' \preceq v_2}$
and $\under{H}{v \preceq v'}$.
\end{definition}

The function \C{lca} used in the premise of \rulelabel{E-Pull-Wait}
computes the LCA of two branches, and assumes it to be unique.  But,
observe that the definition of LCA doesn't guarantee its uniqueness;
it has to be enforced explicitly.  The semantics enforces the
uniqueness of LCA by constraining the branching structure via the
\rulelabel{E-Pull-Wait} rule. The rule merges version $v'$ into
version $v$ only if they are safely mergeable ($\under{H}{v' \mbleto
  v}$), i.e., they have a single LCA, and merging both versions does
not lead to a case where a branch (i.e., its latest version) has
multiple LCAs with remaining branches, and hence gets
\emph{stuck}. Such nuances of LCA, and the guarantees provided by the
system notwithstanding these nuances, are discussed in 
Sec.~\ref{sec:meta}.

\input{mergeable-types}

\input{meta.tex}
